---
title: "Data Sources"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
   
---


```{r include=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(rvest)
library(readxl)

knitr::opts_chunk$set(
	echo = TRUE,
	message=FALSE,
	warning = FALSE,
	results = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%")
```
## Source 1: Abortion Clinic Distances 

Since the overturn of Roe v Wade we believed it would be useful to identify how distances to the nearest abortion provider within each state changed between the overturn in June to July 2022. This data was provided by the [Myers Abortion Facility Database](https://osf.io/pfxq3/) and is public information published on the Open Science Framework website. The database uses public data to as an estimate of travel time and distance for an individual to reach an abortion resource, whether this is a private physician, hospital or a clinic. The distances are calculated at a county level for each month. This information was collected by Caitlin Myers and has information from 2009 to October 1st, 2022.  

## Source 2: Abortion Laws by State 

We are also interested in the abortion laws and restrictions placed in each state. This data comes from [Temple University's Policy Surveillance Program](https://lawatlas.org/datasets/abortion-bans). The data set looks at abortion regulations starting from December 1, 2018 through October 1, 2021. We are also using data on abortion laws restriction status by state after Roe v. Wade was overturned. This data that include updates through November 2022 comes from the non-profit Lucy Burns Institute [Ballotpedia](https://ballotpedia.org/Abortion_regulations_by_state). The data set was not available to download so we scraped it. In addition to abortion laws this source also includes laws defining restrictions on abortions such as state cases which prohibit abortion after specific gestational limits or number of weeks post-fertilization.   

```{r}
bans2018 <-
  read_excel("datasets/abortion_bans_data_Oct 2021.xlsx", sheet = "Statistical Data", range = "A1:H127") %>% 
  janitor::clean_names() %>% 
  rename(
    state = jurisdiction,
    abstatus = prohibit_req) %>% 
  mutate(
    state_abv = state.abb[match(state, state.name)],
    state_abv = ifelse(is.na(state_abv), "DC", state_abv),
    year = lubridate::year(effective_date),
    abstatus = replace(abstatus, abstatus == ".", "0")) %>% 
    select(state, state_abv, year, abstatus) %>% 
  filter(year == 2018)


bans_link <- "https://ballotpedia.org/Abortion_regulations_by_state"
bans_page <- read_html(bans_link)

bans2022 <- 
  bans_page %>%  html_nodes("table") %>% .[2] %>% 
  html_table() %>% .[[1]] %>% 
  janitor::clean_names() %>% 
  rename(
    state = state_abortion_restrictions_based_on_stage_of_pregnancy,
    abstatus = state_abortion_restrictions_based_on_stage_of_pregnancy_2,
    threshold = state_abortion_restrictions_based_on_stage_of_pregnancy_3) %>% 
  mutate(
  state = replace(state, state == "Washington, D.C.", "District of Columbia"),
  state_abv = state.abb[match(state, state.name)],
  state_abv = ifelse(is.na(state_abv), "DC", state_abv),
  year = 2022,
  abstatus = recode(abstatus, "Yes" = "1", "No" = "0")) %>% 
  select(- threshold) %>% 
  slice(-1) %>% 
  head(51) %>% 
  arrange(state) 
```

We then merged both 2018 and 2022 data sets after harmonizing the variable names. The resulting data set was saved as a CSV file for later use in further data processing. The new data set include 4 variables, some of which are listed below:

* `state`: State name
* `abstatus18`: Abortion restriction status in 2018 ( 1 = restricted, 0 = not restricted)
* `abstatus22`: Abortion restriction status in 2022 ( 1 = restricted, 0 = not restricted)

```{r}
abortion_bans <-
  full_join(bans2018, bans2022)%>% 
  arrange(state, year) %>% 
  pivot_wider(
    names_from = year,
    values_from = abstatus) %>% 
  rename(abstatus18 = '2018', abstatus22 = '2022') 

write_csv(abortion_bans, "data_cleaned/abortion_bans_final.csv")
```

## Source 3: Voter Turnout Data 

The voter turnout data was pulled from [US Elections Project](https://www.electproject.org/2022g). This source was created by a Professor in the Department of Political Science at the University of Florida.  

```{r}
library(tidyverse)
library(tidyr)
library(readxl)

voting2022_df =
  read_xlsx("datasets/2022 November General Election.xlsx", sheet = "Turnout Rates", range = "A2:N54") %>%
  janitor::clean_names() %>% 
  rename(state = x1, turnout_estimate2022 = preliminary_total_turnout_estimate, turnoutRate2022 = preliminary_turnout_rate, voting_eligible_pop2022 = voting_eligible_population_vep, voting_age_pop2022 = voting_age_population_vap) %>% 
  select(state, turnout_estimate2022, turnoutRate2022, voting_eligible_pop2022, voting_age_pop2022, state_abv)

voting2018_df = 
    read_xlsx("datasets/2018 November General Election.xlsx", sheet = "Turnout Rates", range = "A2:P54") %>%
  janitor::clean_names() %>%  
  rename(state = x1, turnout_estimate2018 = estimated_or_actual_2018_total_ballots_counted, turnoutRate2018 = estimated_or_actual_2018_total_ballots_counted_vep_turnout_rate, voting_eligible_pop2018 = voting_eligible_population_vep, voting_age_pop2018 = voting_age_population_vap) %>% 
  select(state, turnout_estimate2018, turnoutRate2018, voting_eligible_pop2018, voting_age_pop2018, state_abv)
```

The 2018 and 2022 datasets included TurnoutRate, which is the percent voter turnout in each state based on the number of people that are eligible to vote within that state. The other relevant variables that were kept were the turnout_estimate (the number of people who voted), voting_eligible_pop (the number of people who are eligible to vote), and voting_age_pop (the number of people old enough to vote). The two datasets were merged and difference variables were created to show the difference in each of the variables between the two years. 

```{r}
voting_df = 
  full_join(voting2022_df, voting2018_df, by = "state_abv") %>% 
  select(-state.x) %>% 
  mutate(turnout_estimate_difference = turnout_estimate2022 - turnout_estimate2018, turnout_Rate_difference = turnoutRate2022 - turnoutRate2018, voting_eligible_difference = voting_eligible_pop2022 - voting_eligible_pop2018, voting_age_difference = voting_age_pop2022 - voting_age_pop2018) %>% 
  rename(state = state.y)

voting_df <- voting_df[, c(6, 5, 1, 7, 11, 2, 8, 12, 3, 9, 13, 4, 10, 14)]

write.csv(voting_df,"data_cleaned/Voting Turnout - Final.csv", row.names = FALSE)
```

